Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-09-01 analysis/paper/paper.Rmd
Updated figures and explanation for raman results

All measurements were performed at the Environmental Archaeology Lab at Umeå University using three different instrumentations; NIRS, Raman scattering, and XRF. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was measured twice on different sides, with additional measurements in cases where sampling was difficult. The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that likely were made from the same material, however, for the smaller sites there still remains some uncertainties.  
Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset (Geladi and Linderholm, 2020). PCA modelling was performed using the Prediktera Evince software (www.prediktera.com), which uses the NIPALS algorithm [@bertils19 pp.9-10].

@@

## Near Infrared
NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum. Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups. The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water (REF). 
NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe. The ASD features a detection range of 350 – 2500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background. 
Two parts of the NIR spectra was analysed separately, the visual region (400 – 750 nm) and the 1 000 – 2 500 nm region. In both analyses the data was mean-centred before PCA modelling.

@@

## Raman
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01]. However, where NIRS generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01]. 
The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 – 2 500 cm-1 ^[Instead of reporting Raman measurements in spectral wavelength (nm) they are typically given in wavenumber (cm-1), referred to as *Raman shift*] with a resolution of 9.5 cm-1. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå. 
All Raman spectra were baseline corrected using modified polynomial fitting in R [@Rteam17] using the ChemoSpec package [@hanson16]. The spectra were then mean-centred and normalized using Standard Normal Variate [@shaver01 pp.284-285] in the Prediktera Evince software.

@@

```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark and light groups in the 400 - 750 nm region. Dark material has a higher overall absorbance, with material closer to a white hue feature lower absorbance. The *translucent*/*opaque* group also feature higher absorbance in the ultraviolet region."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_2_average_visnir_spectra.jpg')
```

@@

```{r nir-spectra-comparison, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="Comparison of average spectra for dark, white, and translucent materials. Brecciated quartz is included as it is typically translucent with dark particle inclusion, as a comparison with the translucent group that mainly is made up of seemingly pure quarts."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/nir_spectra_comparison.jpg')
```

@@

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca. 695 cm-1 and 795 cm-1. Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks turned up in measurements collected on shale from XXXX. There is a fairly large band around 500 cm-1 which can be indicative of the presence of amorphous carbon (aC) [@hodkiewicz10], and it is likely the features observed above are another indication of carbon present within the samples. No other features were pr

>>>>

All measurements were performed at the Environmental Archaeology Lab at Umeå University using three different instrumentations; NIRS, Raman scattering, and XRF. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was measured twice on different sides, with additional measurements in cases where sampling was difficult. The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that likely were made from the same material, however, for the smaller sites there still remains some uncertainties.

Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset (Geladi and Linderholm, 2020). PCA modelling was performed using the Prediktera Evince software (www.prediktera.com), which uses the NIPALS algorithm [@bertils19 pp.9-10].

@@

## Near Infrared
NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum. Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups. The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water [@hunt77 p.508].

NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe. The ASD features a detection range of 350 – 2500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background and the collected measurements were averaged for each artefact. Two parts of the NIR spectra were analysed separately, the visual region (400 – 750 nm) and the 1 000 – 2 500 nm region. In both analyses the data was mean-centred before PCA modelling.

@@

## Raman
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01]. However, where NIRS generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 – 2 500 cm-1 ^[Instead of reporting Raman measurements in spectral wavelength (nm) they are typically given in wavenumber (cm-1), referred to as *Raman shift*] with a resolution of 9.5 cm-1. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.

All spectra were averaged for each artefact and the resulting spectra were then baseline corrected using modified polynomial fitting in R [@Rteam17] using the ChemoSpec package [@hanson16]. The spectra were mean-centred and normalized using Standard Normal Variate [@shaver01 pp.284-285] before PCA in the Prediktera Evince software.

@@

```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark (N1), translucent (MV) and white (N1) groups in the 400 - 750 nm region. Dark material has a higher overall absorbance, whereas white and translucent material feature lower absorbance. Brecciated quartz, which is typically described as a translucent material, falls somewhere inbetween the above groups."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_2_breccie_averaged_visnir_spectra.jpg')
```

@@

@@

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca. 695 cm-1 and 795 cm-1. Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks turned up in measurements collected on different types of shale, as well as samples of diabase, apatite, orthoclase, and blue quartz. Studies of clay minerals [@kloprogge17]
have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group. Considering the diagenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.
